---
title: "ESM567 HW#2"
author: "Bernard Romey"
date: "Friday, January 23, 2015"
output:
  word_document:
    fig_caption: yes
  pdf_document: default
---

In order to evaluate the environmental data from the Floriday Everglades for principal components analysis we need to follow the next six steps:

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=6, fig.height=6, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
```

```{r dta, echo=FALSE}
dta <- read.csv("evenv.csv")
env <- na.omit(dta)
rm(dta)
env <- env[,-1]
```

**1.  Use ‘cor.matrix’ function to examine pair-wise relationships among all numerical variables.**

```{r cor.matrix, echo=FALSE}
cor.matrix<-function(x,data=NA,cor.method="pearson",add1to1=F,...){
  # panel.hist function adds the histogram
options(warning=F)
    panel.hist <- function(x)
  {
      usr <- par("usr")
      on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE,breaks=30)
      breaks <- h$breaks
      nB <- length(breaks)
      y <- h$counts
      y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col="lightblue")
      box()
  }
  panel.cor <- function(x, y, digits=2, prefix="", cex.cor)
  {
      usr <- par("usr")
      on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- cor(x, y,method="spearman")
      txt <- format(c(r, 0.123456789), digits=digits)[1]
      txt <- paste(prefix, txt, sep="")
      if(missing(cex.cor)) cex <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex * abs(r))
  }
  panel.smooth2<-function(x, y, bg = NA, pch = par("pch"),cex = 1, col.smooth = "red", span = 2/3,iter = 3,plot1to1=add1to1)
  {
    points(x, y, pch = pch, cex = cex,...)
    ok <- is.finite(x) & is.finite(y)
    if (any(ok)){
        lines(stats::lowess(x[ok], y[ok], f = span, iter = iter), col = col.smooth)
        }
        if(plot1to1){
          mn<-min(c(x,y),na.rm=T)
          mx<-max(c(x,y),na.rm=T)
          new.x<-seq(mn,mx,length.out=10)
          points(new.x,new.x,type='l',col="blue",lwd=2)
        }
  }
  if (class(x)=="formula"){
    x<-model.frame(x,data=data)

  }
  pairs(x,upper.panel=panel.smooth2,lower.panel=panel.cor,diag.panel=panel.hist,
        cex.labels = 1, font.labels=2)
  options(warning=T)
}
```

```{r pair-wise, echo=FALSE, eval=FALSE}
cor.matrix(env) 
```

**2.	Calculate a variance and covariance matrix with a scaled dataset (standardization).**

```{r corr.mx, results='asis', eval=FALSE}
#scaled data
round(var(scale(env)),2) # correlation matrix with log transformed data
kable((cov), digits=2) #Formats table for output
```

**3.	Run PCA in R using “princomp” in the MASS package with the scaled dataset.**

```{r PC,echo=F, results='hide'}
require(MASS)
pca1 <- princomp(env, cor=TRUE) #run PCA on correlatin matrix and save all outputs in 'pca1'
summary(pca)
```

**4.	Run “broken.stick” function to decide how many PCs you should interpret.**

```{r brkStick, echo=FALSE}
broken.stick <- function(p)
  # Compute the expected values of the broken-stick distribution for 'p' pieces.
  # Example: broken.stick.out.20 = broken.stick(20)
  #             Pierre Legendre, April 2007
{
  result = matrix(0,p,2)
  colnames(result) = c("j","E(j)")
  for(j in 1:p) {
    E = 0
    for(x in j:p) E = E+(1/x)
    result[j,1] = j
    result[j,2] = E/p
  }
  return(result)
}
```

```{r Bstick, results='asis', results='hide'}
kable(broken.stick(15))
```

**5.	Once you decide how many PCs you will interpret, report the followings:**

* Percent variance explained by the PC axes

* Eigenvectors of the PC axes

* Scree plot and biplot

**6.  The final step is to interpret the PCA biplot.  You need to use both exploratory results (task 1 and 2) and PCA outputs (e.g., eigenvalues and eigenvectors) to aid your interpretations.  Specifically, you need to interpret the followings:**

* How effectively PCA reduces the original data dimensionality

* Major patterns of the sampling sites in the PCA space based on measured variables.  In other words, what each PC means in terms of measured variables.

------------
For this Principal Component Analysis (PCA) I used the R statistical programming language.

The first step in PCA is to create the variance/Covariance matrix and compare the variances along the primary axis.  The between varaible covariances should also be compared to see where the largest covariances exist.  Standardizing the covariance matrix is recomended if the units of measure between vaiables is not the same, and allows a better comparison between vairables by using the Person Correaltion coefficient:

$PCC=$

Once I decided to use the correlation matrix for analysis, I then used the "pcomp()" function in R to create the cononical matrix of eigenvalues.  The eigenvalues in the cononical matrix are the representation of how much total variance contribution occurs from all variables toward the new components.  


I then used the broken stick criterion to evaluate the importance of the newly created principal components.  Observed eighenvalues greater than the broken stick eigenvalues are considered meaningfull and are thus used for interpretation.  After comparsion, only component 1 and 2 were greater than the broken stick eigenvalues and were selected for interpretation.

**Percent variance explained by the PC axes**

Proportion of Variance (PV) is observed in the eigenvalues in the cononical matrix and is calculated with the following formula:

$PV=\frac{\lambda_i}{\lambda_i}$ where $i=1...p$ variables.  Equivalent to $R^2$ in regression.

Componenet 1 represents 26.3% of the variability along the primary axis,  while  component 2 represents 23.8% of the variation along the secondary component axis (Figure 1).  Thus, both component one and two comprize 50.1% of the variability observed in the data.

**Eigenvectors of the PC axes**

```{r load, tidy=TRUE, results='asis'}
ld <-round(loadings(pca)[,c(1:2)],2) #Loading for PC1 & 2 only
kable(ld)
```

The eigenvector matrix is a relative representaiton of the correlations between variables.

**Scree plot and biplot**
  
```{r Scree, fig.cap="Scree plot showing the variance (eigenvalues) for first 10 principal component."}
plot(pca, main="Scree Plot") #Scree plot
```


```{r Biplot, fig.cap="Biplot of principal component 1 & 2 with"}
biplot(pca, main = "Biplot", xlab = "Comp.1 (26.3%)", ylab = "Comp.2 (23.8%)")
```

6.	Interpret the PCA biplot.  You need to use both exploratory results (task 1 and 2) and PCA outputs (e.g., eigenvalues and eigenvectors) to aid your interpretations.  Specifically, you need to interpret the followings:

**How effectively PCA reduces the original data dimensionality**
  
Effectively reduces fifteen variables down to two synthetic principal components. 

**Major patterns of the sampling sites in the PCA space based on measured variables.  In other words, what each PC means in terms of measured variables.**

*What is the differences between the two synthetic components*

For interpretation, I used a principal component loading level greater than $\mid{.32}\mid$ for a comparative benchmark.

For PC1, low values of Total phophorus concentration in sediments (TPSed) is associated with low values of Total phosphorus concentration in macrophytes (TPeleo), and low values of Total carbon concentration in sediments  (TCSed).  For PC2, high values of Unfiltered Total nitrogen in water column (UNFILT.TN) is associated with high values of Ortho-phosphorus concentration in water column (FILT.PO4) and Silica concentration in water column (SIO2), and low values of Total nitrogen concentration in sediments (TNSed).

The variables seem to be organized into groups.  TCsed, TNeleo, and NTsed are positively correlated.  TPeleo, TPsed, and UNFILT.TP are positivley correlated.  The large group of AFDMsed, CHLAsed, FILT.TP, SIO2, FILT.PO4 are positively correlated, and UNFILT.TN adn FILT.TN are positivley correlated.

Since no single variable has a loading more than 0.45 for both components, the varibles do a poor job representing the new components.

Since the Total Phospherous (TP) are the dominant loading on Principal component one, we will call PC1 the TP gradient.  In addition, Total Nitrogen (TN) is dominant on PC2 so we will call it the TN cradient.
