---
title: "ESM567 HW#2"
author: "Bernard Romey"
date: "Friday, January 23, 2015"
output:
  pdf_document: default
  word_document:
    fig_caption: yes
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=5, fig.height=3.5, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
```

```{r dta, echo=FALSE}
dta <- read.csv("evenv.csv")
dta <- na.omit(dta)
library(dplyr)
dta <- select(dta,-Site)
dta <- scale(dta)
```

> 1.  Use ‘cor.matrix’ function to examine pair-wise relationships among all numerical variables.

```{r cor.matrix, echo=FALSE}
cor.matrix<-function(x,data=NA,cor.method="pearson",add1to1=F,...){
  # panel.hist function adds the histogram
options(warning=F)
    panel.hist <- function(x)
  {
      usr <- par("usr")
      on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE,breaks=30)
      breaks <- h$breaks
      nB <- length(breaks)
      y <- h$counts
      y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col="lightblue")
      box()
  }
  panel.cor <- function(x, y, digits=2, prefix="", cex.cor)
  {
      usr <- par("usr")
      on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- cor(x, y,method="spearman")
      txt <- format(c(r, 0.123456789), digits=digits)[1]
      txt <- paste(prefix, txt, sep="")
      if(missing(cex.cor)) cex <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex * abs(r))
  }
  panel.smooth2<-function(x, y, bg = NA, pch = par("pch"),cex = 1, col.smooth = "red", span = 2/3,iter = 3,plot1to1=add1to1)
  {
    points(x, y, pch = pch, cex = cex,...)
    ok <- is.finite(x) & is.finite(y)
    if (any(ok)){
        lines(stats::lowess(x[ok], y[ok], f = span, iter = iter), col = col.smooth)
        }
        if(plot1to1){
          mn<-min(c(x,y),na.rm=T)
          mx<-max(c(x,y),na.rm=T)
          new.x<-seq(mn,mx,length.out=10)
          points(new.x,new.x,type='l',col="blue",lwd=2)
        }
  }
  if (class(x)=="formula"){
    x<-model.frame(x,data=data)

  }
  pairs(x,upper.panel=panel.smooth2,lower.panel=panel.cor,diag.panel=panel.hist,
        cex.labels = 1, font.labels=2)
  options(warning=T)
}
```

```{r pair-wise, echo=FALSE, eval=FALSE}
cor.matrix(dta[,c(2:15)])  #As a guide, only show 5 variables to check each variable's distribution and pair-wise relationships
```


> 2.	Calculate a variance and covariance matrix with a scaled dataset (standardization).

```{r kable, results='asis', eval=FALSE}
#scaled data
dta1 <-round(var(log(dta+1)),2) #variance/covariance matrix

kable((dta1), digits=1)
```


> 3.	Run PCA using “princomp” in the MASS package with the scaled dataset.

```{r scale,echo=TRUE,results='asis'}
require(MASS)
pca1<-princomp(scale(dta[,-c(3,9)],scale=F)) #run PCA on covariance and save all outputs in 'pca1'


```


> 4.	Run “broken.stick” function to decide how many PCs you should interpret.

```{r brkStick, echo=FALSE}
broken.stick <- function(p)
  # Compute the expected values of the broken-stick distribution for 'p' pieces.
  # Example: broken.stick.out.20 = broken.stick(20)
  #             Pierre Legendre, April 2007
{
  result = matrix(0,p,2)
  colnames(result) = c("j","E(j)")
  for(j in 1:p) {
    E = 0
    for(x in j:p) E = E+(1/x)
    result[j,1] = j
    result[j,2] = E/p
  }
  return(result)
}
```

```{r kable2, results='asis'}

kable(broken.stick(15))
```


> 5.	Once you decide how many PCs you will interpret, report the followings:
  
  •	Percent variance explained by the PC axes


  •	Eigenvectors of the PC axes


  •	Scree plot and biplot
  
  

> 6.	Interpret the PCA biplot.  You need to use both exploratory results (task 1 and 2) and PCA outputs (e.g., eigenvalues and eigenvectors) to aid your interpretations.  Specifically, you need to interpret the followings:

  •	How effectively PCA reduces the original data dimensionality 


  •	Major patterns of the sampling sites in the PCA space based on measured variables.  In other words, what each PC means in terms of measured variables.



